Value-Suppressing Uncertainty Maps

### 2017-12-22 paper/main.tex
Acknowledgments
Probably should add this section before I forget.
Added missing cite
More discussion writing.

When selecting a visual channel to represent uncertainty, we recommend channels with both a strong semiotic connection to uncertainty (such as those recommended by MacEachren et al.~\cite{maceachren2012visual}) as well as a relatively large number of perceptually distinguishable levels (such as those evaluated by Boukhelifa et al. \cite{boukhelifa2012evaluating}). Here, we focus on a combination of increasing luminance and decreasing saturation. Correll \& Gleicher~\cite{correll2014error} provide evidence that even audiences without statistical training can successfully interpret uncertainty information encoded in these channels.

@@

Our experimental results have a number of implications for designers wishing to employ bivariate maps for encoding uncertainty. Both of our experiments show that the way that uncertainty information is presented can have a measurable impact of decision-making. VSUPs make a number of design choices in presenting uncertainty. Firstly, they present uncertainty information simultaneously (superimposed) with value, rather than in juxtaposed charts. Secondly, they encode value and uncertainty in discrete, quantized bins, rather than continuously. Lastly, they alias and suppress uncertain, rather than certain, values. Our experiments provide specific performance information about the effects of each of these choices on information fusion tasks.

@@

Gleicher et al.~\cite{gleicher2011visual} propose three major categories for visualizations that support comparison: juxtaposition (where two variables are visualized in adjacent charts), superposition (where the two variables are encoded simultaneously in a single chart) and explicit encoding (where the two variables are reduced to a single calculated variable, which is directly visualized). Juxtaposition has a number of advantages for bivariate data of this sort: firstly, there are no issues of the integrality of different components of color perception, as there are separate maps for both uncertainty and value. Secondly, juxtaposed displays allow viewers to separately consider patterns in value and uncertainty.

\subsection{Discrete vs Continuous Encodings}

When using color to encode data, designers must make the decision whether or not to separate the data into discrete classes of color outputs, or to employ a continuous and class-less scale. This, in turn, introduces a trade-off between \emph{quantization error} (the sampling error introduced by binning the data) and \emph{perceptual error} (the error introduced by the viewer as they decode the color). For bivariate color maps, where often the color properties used to encode value and uncertainty are not perceptually separable, this perceptual error can be quite high. We observed this high error in our identification task, where continuous maps, despite, in theory, having greater expressivity in encoding color and uncertainty, had significantly worse accuracy than the discretized bivariate maps. This is despite the fact that we had comparatively few categories (16 in total).

@@

We designed VSUPs to promote risk-averse decision-making. In the limit of uncertainty VSUPs have only one output color, and so explicitly prevent any decoding of value. This narrowing of options is reflected not only in the quantization scheme, but also in the shape of the legend, narrowing as the uncertainty increases. 
%\sizeFig

@@

>>>>

When selecting a visual channel to represent uncertainty, we recommend channels with both a strong semiotic connection to uncertainty (such as those recommended by MacEachren et al.~\cite{maceachren2012visual}) as well as a relatively large number of perceptually distinguishable levels (such as those evaluated by Boukhelifa et al. \cite{boukhelifa2012evaluating}). Here, we focus on a combination of increasing luminance and decreasing saturation, as in Hengl \& Toomanian~\cite{hengl2006maps}. Correll \& Gleicher~\cite{correll2014error} provide evidence that even audiences without statistical training can successfully interpret uncertainty information encoded in these channels.

@@

Our experimental results have a number of implications for designers wishing to employ bivariate maps for encoding uncertainty. Both of our experiments show that the way that uncertainty information is presented can have a measurable impact on decision-making. VSUPs make a number of design choices in presenting uncertainty. Firstly, they present uncertainty information simultaneously (superimposed) with value, rather than in juxtaposed charts. Secondly, they encode value and uncertainty via discrete, quantized bins, rather than continuously. Lastly, they are designed to alias and suppress uncertain values, rather than certain values. Our experiments provide evidence about the potential impact of each of these choices on performance for information fusion tasks. We will discuss the impact of our experimental findings on the consideration of each of these design decisions in turn.

@@

Gleicher et al.~\cite{gleicher2011visual} propose three major categories for visualizations that support comparison: juxtaposition (where two variables are visualized in adjacent charts), superposition (where the two variables are encoded simultaneously in a single chart) and explicit encoding (where the two variables are reduced to a single calculated variable, which is directly visualized). Juxtaposition has a number of advantages for bivariate data of this sort: firstly, there are no issues of the integrality of different components of color perception, as there are separate maps for both uncertainty and value. Secondly, juxtaposed displays allow viewers to separately consider patterns in value and uncertainty. The literature also suggests that superposition, which requires a bivariate scale, results in poor accuracy even for basic identification tasks~\cite{wainer1980empirical}.

However, our results indicate that juxtaposed visualizations have a substantial drawback for information fusion tasks.

\subsection{Discrete vs. Continuous Encodings}

When using color to encode data, designers must make the decision whether or not to separate the data into discrete classes of color outputs, or to employ a continuous and class-less scale. This, in turn, introduces a trade-off between \emph{quantization error} (the sampling error introduced by binning the data) and \emph{perceptual error} (the error introduced by the viewer as they decode the color). There is also a cost in terms of expressivity and consistency: a designer might wish for quantitative, continuous information to be represented continuously~\cite{mackinlay1986automating}.

However, for bivariate color maps, where often the color properties used to encode value and uncertainty are not perceptually separable, error in decoding values can be quite high.  In our identification experiment, this perceptual error was high. Continuous maps performed the worst of all of our other bivariate map types (47\% vs. 63\% accuracy for discrete maps). This is likely due to the non-linearity of

@@

We designed VSUPs to promote risk-averse decision-making. In the limit of uncertainty VSUPs have only one output color, and so explicitly prevent any decoding of value. This narrowing of options is reflected not only in the quantization scheme, but also in the shape of the legend, narrowing as the uncertainty increases. This is despite the fact that the continuous maps were constructed via interpolations in CIELAB colorspace, which was designed for perceptual uniformity~\cite{mahy1994evaluation}. 
%\sizeFig

@@

\section{Acknowledgments}

This work was supported by a Moore Foundation Data-Driven Discovery Investigator Award.
